---
title: "Practical Machine Learning Prediction Assignment Writeup"
author: "Nishay Naser Pasha"
output: html_document
---

###Introduction

The goal of this project is to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants to predict the manner in which they did the exercise.

###Loading Packages

```{r loadingPackages, warning=FALSE, message=FALSE}
library(caret)
library(rpart)
library(rattle)
library(rpart.plot)
library(randomForest)
library(RColorBrewer)
```

###Setting the Seed

The seed is set to be able to reproduce the results.

```{r setSeed}
set.seed(12345)
```

###Loading the Data

The data is loaded.

```{r loadingData}
trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
training <- read.csv(url(trainUrl), na.strings=c("NA","#DIV/0!",""))
testing <- read.csv(url(testUrl), na.strings=c("NA","#DIV/0!",""))
```

###Partitioning the Data

60% of the data is partitioned as the Training Set and 40% is used as the Testing Set.

```{r partitioningData}
inTrain <- createDataPartition(y=training$classe, p=0.6, list=FALSE)
myTraining <- training[inTrain, ]; myTesting <- training[-inTrain, ]
dim(myTraining)
dim(myTesting)
```

###Cleaning the Data

```{r cleaningData}
myDataNZV <- nearZeroVar(myTraining, saveMetrics=TRUE)
myNZVvars <- names(myTraining) %in% c("new_window", "kurtosis_roll_belt", "kurtosis_picth_belt",
"kurtosis_yaw_belt", "skewness_roll_belt", "skewness_roll_belt.1", "skewness_yaw_belt",
"max_yaw_belt", "min_yaw_belt", "amplitude_yaw_belt", "avg_roll_arm", "stddev_roll_arm",
"var_roll_arm", "avg_pitch_arm", "stddev_pitch_arm", "var_pitch_arm", "avg_yaw_arm",
"stddev_yaw_arm", "var_yaw_arm", "kurtosis_roll_arm", "kurtosis_picth_arm",
"kurtosis_yaw_arm", "skewness_roll_arm", "skewness_pitch_arm", "skewness_yaw_arm",
"max_roll_arm", "min_roll_arm", "min_pitch_arm", "amplitude_roll_arm", "amplitude_pitch_arm",
"kurtosis_roll_dumbbell", "kurtosis_picth_dumbbell", "kurtosis_yaw_dumbbell", "skewness_roll_dumbbell",
"skewness_pitch_dumbbell", "skewness_yaw_dumbbell", "max_yaw_dumbbell", "min_yaw_dumbbell",
"amplitude_yaw_dumbbell", "kurtosis_roll_forearm", "kurtosis_picth_forearm", "kurtosis_yaw_forearm",
"skewness_roll_forearm", "skewness_pitch_forearm", "skewness_yaw_forearm", "max_roll_forearm",
"max_yaw_forearm", "min_roll_forearm", "min_yaw_forearm", "amplitude_roll_forearm",
"amplitude_yaw_forearm", "avg_roll_forearm", "stddev_roll_forearm", "var_roll_forearm",
"avg_pitch_forearm", "stddev_pitch_forearm", "var_pitch_forearm", "avg_yaw_forearm",
"stddev_yaw_forearm", "var_yaw_forearm")
myTraining <- myTraining[!myNZVvars]
# Review the revised value of observations
dim(myTraining)
```

The first column of ID is removed to ensure that the Machine Learning Algortihm can run smoothly

```{r removeID}
myTraining <- myTraining[c(-1)]
```

Updating the NA values

``` {r updatingNA}
trainingV3 <- myTraining 
for(i in 1:length(myTraining)) { 
if( sum( is.na( myTraining[, i] ) ) /nrow(myTraining) >= .6 ) { 
for(j in 1:length(trainingV3)) {
if( length( grep(names(myTraining[i]), names(trainingV3)[j]) ) ==1)  { 
trainingV3 <- trainingV3[ , -j] #Remove that column
}}}}
dim(trainingV3)
```

``` {r}
myTraining <- trainingV3
rm(trainingV3)
```

Completing the same transformations on the Testing Set.

``` {r}
clean1 <- colnames(myTraining)
clean2 <- colnames(myTraining[, -58])
myTesting <- myTesting[clean1]
testing <- testing[clean2]
dim(myTesting)
```

``` {r}
dim(testing)
```

###Coercing Data into the Same Type

``` {r sameType}
for (i in 1:length(testing) ) {
for(j in 1:length(myTraining)) {
if( length( grep(names(myTraining[i]), names(testing)[j]) ) ==1)  {
class(testing[j]) <- class(myTraining[i])
}}}
testing <- rbind(myTraining[2, -58] , testing)
testing <- testing[-1,]
```

###Development of Decision Tree Machine Learning Algorithm for Prediction

``` {r}
modFitA1 <- rpart(classe ~ ., data=myTraining, method="class")
```

Developing the Decision Tree

``` {r decisionTree}
fancyRpartPlot(modFitA1)
```

Prediction Algorithm

``` {r}
predictionsA1 <- predict(modFitA1, myTesting, type = "class")
```

Confusion Matrix

``` {r}
confusionMatrix(predictionsA1, myTesting$classe)
```

###Development of Random Forest Machine Learning Algorithm for Prediction

``` {r}
modFitB1 <- randomForest(classe ~. , data=myTraining)
predictionsB1 <- predict(modFitB1, myTesting, type = "class")
confusionMatrix(predictionsB1, myTesting$classe)
```

###Files for Executing Test Cases

``` {r testCases}
predictionsB2 <- predict(modFitB1, testing, type = "class")
```

Creating the files in the current working directory

``` {r}
pml_write_files = function(x){
n = length(x)
for(i in 1:n){
filename = paste0("problem_id_",i,".txt")
write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
}}
pml_write_files(predictionsB2)
```
